---
title: "bank_customer_acct_exp"
author: "Lujia Deng"
date: "12/12/2020"
output: html_document
---

```{r setup, include=FALSE}
library(reticulate)
use_python("C:/Users/ldeng/Python/Python39", required = TRUE)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{python}
import warnings
warnings.filterwarnings('ignore')

import matplotlib
matplotlib.use("TkAgg")
from matplotlib import pyplot as plt
import pandas as pd
import seaborn as sns
import numpy as np

pd.set_option('display.max_columns', None)
pd.set_option('display.max_rows', None)
pd.set_option('max_colwidth', 100)
```


#### **Task 1:** Explore account characteristics: 1) Whether an account has a credit card or not? 2) Whether an account has a loan or not? 3) The average balance for the account


```{python}
# import data
acct = pd.read_csv("./anly-503-a5/data/accounts_analytical.csv") 
txn = pd.read_csv("./anly-503-a5/data/transactions.csv")

# calculate the average
avg_balance = txn.groupby(by="account_id")['account_id','balance'].agg('mean')

# merge into acct dataset
acct_vis = acct.merge(avg_balance['balance'], how="left", on="account_id")[["credit_cards", "loan_date", "balance"]]

# clean the vis data
acct_vis['credit_cards'] = acct_vis['credit_cards'].apply(lambda x: 0 if pd.isnull(x) else 1)
acct_vis['loan_date'] = acct_vis['loan_date'].apply(lambda x: 0 if pd.isnull(x) else 1)

acct_vis = acct_vis.rename(columns = {'credit_cards': 'has_credit_cards', 'loan_date':'has_loan', 'balance':'avg_balance'})

# Create an account type
acct_vis['type'] = None
acct_vis.loc[(acct_vis['has_credit_cards'] == 0) & (acct_vis['has_loan'] == 0), 'type'] = 'no cards, no loan'
acct_vis.loc[(acct_vis['has_credit_cards'] == 1) & (acct_vis['has_loan'] == 0), 'type'] = 'has cards, no loan'
acct_vis.loc[(acct_vis['has_credit_cards'] == 0) & (acct_vis['has_loan'] == 1), 'type'] = 'no cards, has loan'
acct_vis.loc[(acct_vis['has_credit_cards'] == 1) & (acct_vis['has_loan'] == 1), 'type'] = 'has cards, has loan'

acct_vis.head()
```


```{python}
sns.set_palette(["#2A9D8F", '#E9C46A', '#F4A261', '#E76F51', '#264653'])
plt.figure(figsize=(20, 20))
g = sns.FacetGrid(acct_vis, col="has_credit_cards",  row="has_loan", margin_titles=True, despine=False, hue="type")
g.map_dataframe(sns.histplot, x="avg_balance");
g.set_xlabels('Account Average Balance');
g.set_ylabels('Count', labelpad=1);
#g.set_axis_labels('Account Average Balance', "Count");
g.fig.subplots_adjust(wspace=0, hspace=0)
plt.show()
```




#### **Task 2:** What is the distribution of all loans and what are their characteristics?


```{python}
# get loan data
loan = acct.dropna(subset=['loan_date'])[['loan_date', 'loan_amount', 'loan_payment', 'loan_term', 'loan_status', 'loan_default']]

# correct date data type
loan['loan_date'] = [np.datetime64(x) for x in loan['loan_date']]

# retrieve the year information
loan['loan_year'] = np.array(loan['loan_date']).astype('datetime64[Y]').astype(int) + 1970 

loan.head()
```



```{python message=FALSE, warning=FALSE}

# distribution
fig, ax = plt.subplots(2, 2, figsize = (12,12))
                       
sns.distplot(loan['loan_amount']/3, ax = ax[0,0]);
ax[0,0].set_title('Loan Amount Distribution (in thousands $)')
ax[0,0].set_ylabel("Density")

sns.distplot(loan['loan_term'], ax = ax[0,1]);
ax[0,1].set_title('Loan Term Distribution')
ax[0,1].set_ylabel("Density")

sns.distplot(loan['loan_year'], ax = ax[1,0]);
ax[1,0].set_title('Loan Distribution over Year')
ax[1,0].set_ylabel("Density")

# How many loans are current and expired? What proportions of current and expired loans are default?
default = list(set(loan['loan_default'].tolist()))
df_default = loan.groupby(['loan_status', 'loan_default'])['loan_status'].count().unstack('loan_default')
default_mixture = df_default.plot(kind='bar', stacked=True, legend = True, ax = ax[1,1])

plt.show()
```



**Characteristics of the loans:**


1. The distribution of loan amount is right-skewed, very few loans exceeding 600K dollars


2. Loans are uniformly distributed over loan terms


3. More loans are issued in 1997. Overall, the numebr of loans issued is increasing over time


4. The proportions of loan default in current and expired loans are relatively the same.



#### **Task 3:** Is there a relationship between a good or bad loan and the time between an account is opened an the loan is created? Is there a specific set of accounts that seem to be at higher or lower risk of defaulting?



```{python}

# get accounts with loan
acct = acct.dropna(subset=['loan_date'])

# adjust the data type for datetime columns
acct['loan_date'] = [np.datetime64(x) for x in acct['loan_date']]
acct['acct_creation_date'] = [np.datetime64(x) for x in acct['acct_creation_date']]

# get the time between account creation and loan issued
acct['day_diff'] = acct['loan_date'] - acct['acct_creation_date']
acct['day_diff'] = acct['day_diff'].dt.days

```

##### **1. Is there a relationship between a good or bad loan and the time between an account is open and the loan is created?**

```{python}

fig, ax = plt.subplots(1, 1)
sns.boxplot(x="loan_default", y="day_diff", data=acct, ax = ax)
ax.set_title('Loan Default and Time Difference \n between Account and Loan Creation')
ax.set_ylabel('Time Difference (days)')
ax.set_xlabel('Loan Default')
plt.show()

```

We can observe that, accounts with defaulted loans have shorter days between account creation and loan issuance, compared with accounts with good loans. However, the time difference might not be significant, given that there is a great overlap between two boxplots. 


##### **2. Find a specific set of accounts that seem to be at higher risk of defaulting**


**2.1 What are some characteristics of accounts with defaulted loan?**


I am using a scatter matrix plot with possible relevant attributes to explore the characteristics of accounts with loan default:


```{python}
pair_df = acct[['num_customers', 'loan_amount', 'loan_default',
               'max_withdrawal', 'min_withdrawal', 'max_balance', 'min_balance']]

sns.pairplot(pair_df, hue="loan_default");

plt.show()
```


**From the scatter matrix plot above, I have the following observations about loan default:**

1. Default loans are all associated with single-customer accounts

2. They tend to have higher amounts for max_withdrawl (max_withdrawal : loan_amount ratio equals or above 1)

3. They are also more likely to have lower amounts (below 20, close to 0) for min_withdrawal

Therefore, I am going to pull out all the accounts with current loans and have the above characteristics.


```{python}
# get all accounts with current loans
current_loan = acct[acct['loan_status'] == 'current']

# calculate max_withdrawal/loan_amount ratio
current_loan['max_withdrawal_ratio'] = current_loan['max_withdrawal']/current_loan['loan_amount']
# Get all current loans with Characteristics 1 and 2

risky = current_loan[current_loan['num_customers']==1]
risky = risky[risky['max_withdrawal_ratio'] >= 1]
risky = risky[risky['min_withdrawal'] <= 20]

risky[['account_id', 'num_customers', 'loan_status', 'max_withdrawal_ratio', 'min_withdrawal']]
```


#### **Task 4:** For the account with the highest number of transactions, make a time series line plot for the behavior of the account over time, including all debits and credits, the different methods, and the with the different categories.


```{r, message=FALSE}
library(tidyverse)

# read in data
acct = read.csv("./anly-503-a5/data/accounts_analytical.csv") 
txn = read.csv("./anly-503-a5/data/transactions.csv")
```


##### **1. Preprocess the Data**



**2.1 Which account has the highest number of transactions?**

```{r}
# get the account with the highest number of transactions

count_txn <- txn %>% select(account_id, id) %>% group_by(account_id) %>% summarise(count = n_distinct(id)) %>% rename(num_txn = count)

max_num_txn_acct <- count_txn %>% filter(num_txn == max(num_txn)) %>% select(account_id)

print(paste("The account with the highest number of transactions is", max_num_txn_acct))
```


**2.2 Get transaction history for this account and prepare the data for plotting**


```{r}

# get transaction info for this account
txn_hist <- txn %>% filter(account_id == max_num_txn_acct[[1]])

# correct data type and group dates by year & month
txn_hist$date <-as.Date(txn_hist$date)

txn_hist$year_month = (format(txn_hist$date,'%Y-%m'))
txn_hist$year_month = as.Date(paste(txn_hist$year_month,"-01",sep=""))


# Assign category blank as 'others'
temp_category <- as.character(txn_hist$category) 
temp_category[temp_category == ""] <- 'others'
txn_hist$category <- as.factor(temp_category)


# Assign type blank as 'others'
temp_method <- as.character(txn_hist$method) 
temp_method[temp_method == ""] <- 'others'
txn_hist$method <- as.factor(temp_method)


# create a new df for ploting by year_month

df_year_month <- txn_hist %>% select(type, amount, method, category, year_month) %>% group_by(type, method, category, year_month) %>% summarise(sum(amount)) %>% rename('amount' = 'sum(amount)')

summary(df_year_month)
```




##### **3. Make the Multi-Panel Time Series Plot**



```{r}
# create the plot
txn_hist_lineplot <- ggplot(df_year_month, aes(x=year_month, y=amount)) + geom_line(aes(color = method, linetype = type), size = 1, alpha = 0.5) + scale_color_manual(values = c("#00AFBB", "#E7B800", "darkgreen"))+ xlab("")
txn_hist_lineplot + facet_wrap(.~category, scales = "free_y") + 
  labs(title = "Account 8261: Transaction History", subtitle = "The plot is splitted into multiple panels based on the payment category.\n Two line types indicate whether the payment is a credit or a debit, \n and three color encodes payment methods.") + theme_minimal()
```


